Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-02-10 boettiger.tex
a few updates

\subsubsection*{Hidden assumptions}
%Differing system dynamics
%% OKAY, so this is a bit of a stretch.

The underlying assumption that the system contains a saddle-node bifurcation,
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome --
they are characterized by a steady approach towards the alternative stable state, 
such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977},
with stochasticity introduced from directly through the demographic process.
This model is forced through a bifurcation by gradually increasing the $a$ parameter,
which increases can be thought of as an increasing toxicity of the environment
(from $a_0 = 100$ increasing at constant rate of 0.09 units/unit time).
Other parameters are: $Xo = 730$, $e = 0.5$, $K = 1000$, $h = 200$.
We run this model over a time interval from 0 to 500.
and sample at 40 evenly spaced time points, which were used for subsequent analysis.
We refer to this as the simulation under a degrading environment.
We repeat this simulation with constant value of $a$, and which we refer to as the simulation in a stable environment.
For the convenience of the inquisitive reader, we have also provided a simple function in the
associated R package where the user can vary the sampling scheme and parameter values and rerun this analysis.

@@

Because ROC curves compare areas under the probability distributions they can always be presented on a uniform 0-1 axis
for both true positives and false positives, allowing us to overlay the curves from different data on the same plot,
Figure~\ref{fig:rocdata}.

>>>>

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation,
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome --
they are characterized by a steady approach towards the alternative stable state, 
such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

@@

This corresponds to the well-studied ecosystem model of over-exploitation~\citep{Noy-Meir1975, May1977},
with stochasticity introduced from directly through the demographic process.
This model is forced through a bifurcation by gradually increasing the $a$ parameter,
which increases can be thought of as an increasing toxicity of the environment
(from $a_0 = 100$ increasing at constant rate of 0.09 units/unit time).
Other parameters are: $Xo = 730$, $e = 0.5$, $K = 1000$, $h = 200$.
We run this model over a time interval from 0 to 500
and sample at 40 evenly spaced time points, which were used for subsequent analysis.
We refer to this as the simulation under a degrading environment.
We repeat this simulation with constant value of $a$, and which we refer to as the simulation in a stable environment.
For the convenience of the inquisitive reader, we have also provided a simple function in the
associated R package where the user can vary the sampling scheme and parameter values and rerun this analysis.

@@

%Because ROC curves compare areas under the probability distributions they can always be presented on a uniform 0-1 axis
%for both true positives and false positives, allowing us to overlay the curves from different data on the same plot,
%Figure~\ref{fig:rocdata}.
